project-plan

### 2025-10-02 project-plan.qmd
Adding sources + inserting feedback from coordinator

---
title: "project-plan"
date: today 
bibliography: bibliography.bib
csl: vancouver.csl
format: html
---

<!-- The project plan should be 3-4 pages long and written in
Rmarkdown (like this document). Remove all comments in italic when you 
use this document as a template for your own project plan. -->

<!-- This is an R Markdown document. Markdown is a simple formatting
syntax for authoring HTML, PDF, and MS Word documents. For more
details on using R Markdown see http://rmarkdown.rstudio.com -->

<!-- This is a comment and will not be present in the compiled
document. Try "knitting" this document by clicking the knit button up
to the left, or by running `rmarkdown::render("project-plan.Rmd")` in
the console and you should only see headings -->

<!--NEED TO ADD SOURCES IN TEXT-->

Introduction
============
Injury has reached epidemic proportions worldwide, disproportionately affecting the most productive and the youngest
members of society. Trauma, often synonymously used with serious injury, is defined as a physical injury and the
body's associated response [@Szolnoky2025]. Furthermore, trauma remains a leading cause of death and disability
worldwide, where it is the foremost cause of death among people under the age of 45 and ranks as the fourth leading
cause of death across all ages [@Maerz2009].

@@

Among trauma-related conditions, traumatic brain injury (TBI) is a major cause of death and disability worldwide and
constitutes a significant public health burden (3). TBI is the result of an external force causing neurological
impairments that may be either transient or permanent (4). Furthermore, TBI should not be regarded as a single event
but rather as a dynamic process. The initial trauma often triggers a secondary pathophysiological cascade, including ischemia, oedema, and metabolic disturbances, that develop over time and critically influence long-term outcomes (5). As such, patients with TBI often require admission to the intensive care unit (ICU).

@@

Initial management of TBI spans prehospital and early in-hospital phases and centres on rapid physiological
stabilisation and diagnostics in order for secondary injuries to be limited. Protection of the airway in patients
with impaired consciousness, adequate oxygenation and normocapnia, and fast correction of hypotension are all
prioritised in the initial phase of management as hypoxia and low blood pressures are associated with an increased
mortality (6–9). Non-contrast head CT should be performed without delay to identify surgically remediable lesions.
Depending on mechanism and physiology, whole-body CT may be used selectively, as randomised data show no mortality
advantage over targeted imaging, while observational work underscores the benefit of immediate CT pathways for
severely injured patients (10,11). Haemostatic strategies in the early stages, such as administering tranexamic acid
within three hours of injury, can reduce mortality (12). After this initial phase, depending on the patient's condition, management transitions to other areas of the hospital, where sustained prevention of secondary systemic
and intracranial insults becomes the primary focus.

The management of TBI in the ICU is particularly challenging. Given the heterogeneity and dynamic nature of the
condition, care must prioritise the prevention and limitation of secondary systemic threats such as hypoxia,
hypercapnia, arterial hypotension, hyponatraemia and pyrexia. Equal attention must also be directed toward secondary
intracranial threats, including expanding haematomas, cerebral contusions and rises in intracranial pressure. (5–7)
Although these complexities are often anticipated, they render ICU care especially vulnerable to deviations from
optimal practice, thereby posing risks to both patient safety and overall quality of care.

The intensive care of patients with TBI has been addressed in both guidelines and reviews. Multiple recent reviews
and guideline updates consistently refer to the Brain Trauma Foundation (BTF) Fourth Edition, which remains a central
reference for modern management (5–9). According to the BTF, treatment should be initiated when intracranial pressure (ICP) exceeds 22 mmHg. Cerebral perfusion pressure (CPP) should be maintained between 60–70 mmHg. Prophylactic
hyperventilation should be avoided within the first 24 hours, and corticosteroid usage should be refrained from.

<!-- MGW: Try to find a balance when it comes to the level of detail that you use. The specific ICD and CPP targets feel a bit too specific. -->

@@

Despite the presence of systems and guidelines in place for these patients, trauma care remains vulnerable to errors,
deviations from established practice, and treatment delays, all of which may negatively influence outcomes. In
response, the American College of Surgeons developed the Trauma Quality Improvement Program (TQIP), a risk-adjusted
benchmarking system designed to assess the quality of trauma care <!-- MGW: I would mention ACS TQIP as one example of a TQIP and refer to the WHO guidelines as more general. I suggest that you start by introducing the concept of TQIP by referring to the WHO guidelines and then mention ACS TQIP as one example -->. Through audit filters and peer review, it
identifies opportunities for improvement (OFI) and ultimately reduces trauma-related mortality (10,11). The WHO, however, insist that the trauma quality improvement programmes must be a continuous, multidisciplinary process. This
entails audit filters, structured morbidity and mortality and preventable death case reviews in order to find
opportunities for improvement (OFI). They also discuss that the formulation of corrective strategies or action plans
in response to identified OFI:s is essential to trauma quality improvement programmes. Furthermore, the WHO
emphasises that TQIP must ensure that corrective measures addressing identified OFI result in tangible and verifiable
improvements in outcomes, thereby confirming that deficiencies are not only recognised but also effectively resolved. (Källa WHO) <!-- add WHO source -->

Previous studies and reviews have discussed preventable trauma deaths where recurring patterns occur, namely, delays in imaging and care, missed injuries, deviations in care regarding airway-related care, haemorrhage control, and fluid management (12–16). Importantly, some of these findings were observed in high-performing level I trauma centres, underscoring that even mature systems remain vulnerable to error. Furthermore, injuries to the central nervous system account for a substantial proportion of trauma-related deaths, as demonstrated in analyses from large registries such as the Arkansas Trauma Registry (16). Yet, while previous research has described OFI in trauma resuscitation and in general ICU trauma cohorts, relatively little is known about the prevalence, nature, and determinants of OFI specifically among patients with TBI.

<!-- MGW: I think this paragraph is good, but I suggest adding a subheading like "Rationale for this study" or something like that. -->

TBI remains a challenge within modern trauma systems, not only due to its contribution to mortality but also due to
the magnifying impact on the outcome because of small process deviations. Delays in imaging, inconsistent airway
management, and lapses in haemodynamic measures are all examples that can result in irreversible damage. Although
TQIP have been developed to identify and correct such lapses, with the WHO stressing the importance of linking OFIs
to corrective actions and verification of bettered outcomes, most research has been done on mixed trauma cohorts.
This, on the other hand, leaves TBI-specific vulnerabilities undercharacterised. Recent work by Szolnoky et al.
suggests that, despite overall declines in OFI, patients with TBI remain at risk of preventable shortcomings,
underscoring the need for TBI-specific quality improvement efforts (23).

Aim
=====
This study aims to identify and describe opportunities for improvement in the care of adult trauma patients
with TBI, and to examine patient- and process-related factors associated with their occurrence.

@@

Study Design
-----------
A single-centre, registry-based retrospective cohort study was conducted at Karolinska University Hospital in Solna
(NKS). The trauma registry for NKS is a comprehensive trauma registry that is part of the Swedish National Trauma
Registry, SweTrau, and is compliant with the Utstein template for uniform reporting after trauma (Källa  Utstein
Template). As such, the registry prospectively records patient demographics, prehospital and emergency department
physiology, times of key interventions such as imaging, surgery and transfer, injury severity scores, and in-hospital outcome.

SweTrau includes patients of all ages who have sustained serious injuries from traumatic events
such as falls, traffic incidents, or other external violence. Since 2024<!-- MGW: Has this changed over the years? Otherwise I suggest that you remove the year. -->, patients are included if a trauma
team activation has occurred, regardless of injury severity, or if they have a New Injury Severity Score (NISS)
greater than fifteen without trauma team activation. Patients transferred to a hospital within seven days of the
traumatic event with a NISS greater than fifteen are also included. Exclusion criteria are cases where the only
traumatic condition is chronic subdural haematoma, trauma alarms without an underlying traumatic event, patients with
protected identity, and patients declared dead on arrival. Since these inclusion and exclusion criteria have changed
over time, patients in this study will be filtered to ensure they fulfil the criteria relevant to the study period
(Källa SweTrau årsbok 2024). <!--Use the 2023 version instead? Data used is up to 2023 -->

@@

Setting
---------
Karolinska University Hospital in Solna houses Trauma Centre Karolinska (TCK), Sweden’s largest provider of advanced
trauma care for both adults and children<!-- MGW: Do you have a source for this? -->. The trauma center has around-the-clock availability of key services, including neurosurgery, general
surgery, orthopaedic surgery, anaesthesia and intensive care, as well as diagnostic and interventional radiology and
consultants from other relevant specialities. (Källa   från American College of surgeons, plus källa  som säger NKS
har alla dessa saker). Also, the ICU units in Karolinska Universitets sjukhuset are all recognised as Level 3 units
according to the **Swedish Society of Anaesthesiology and Intensive Care Medicine**. (källa)

<!-- The numbers below belong to the results section, so they should be moved there. -->

Furthermore, the trauma registry includes about 14000 patients admitted between 2012 and 2023. The trauma care
quality database is a subset of this registry, encompassing approximately 8,000 patients selected for review between
2014 and 2023. Screening for potential opportunities for improvement was implemented in 2013 for selected patient
cases. Initially, case selection was based on a limited number of audit filters and the judgment of a small group of
clinicians directly involved in trauma care. Since 2017, a standardised set of audit filters has been consistently
applied, and all trauma patients admitted to the hospital are screened for potential opportunities for improvement.
The process has expanded to include a broader group of senior clinicians and nursing staff, ensuring a systematic and
comprehensive review. Patients flagged through the audit filters, and therefore have potential OFI, undergo
discussion in structured multidisciplinary morbidity and mortality conferences, where representatives from relevant
specialties and professions involved in trauma care are involved.  The discussion focuses on establishing whether
there were any preventable events or deviations from standards of practice. The final classification of whether an
opportunity for improvement is present is determined by consensus among the peer-review group. As such, subjectivity
is minimised and ensures that decisions reflect a collective judgement. Whether or not an OFI is identified, the
decision of the review is recorded in the trauma care quality database, ensuring systematic documentation of positive
findings and cases without deviations.

Participants
------------
All adult patients with TBI recorded in the Karolinska trauma registry between 2012 and 2023 who also had been
screened for opportunity for improvements were included in this study. Furthermore, TBI was defined as a GCS score of
≤8 combined with an Abbreviated Injury Scale ≥3 for the head region. Patients were eligible irrespective of level of
care.

Patients younger than 18 years were excluded, as their trauma follows separate clinical care systems and quality <!-- is this true? could not find source-->
review processes that differ from those for adults . Also excluded were patients declared dead on arrival, since
assessment of opportunities for improvement in these cases is not relevant to the aim of this study.

@@

The primary outcome of this study was the presence of one or more opportunities for improvement during the management
of patients with TBI. As mentioned, OFIs were determined after a screening process with audit filters and a
subsequent review in multidisciplinary morbidity-and-mortality conferences. Following the peer review, the result of
each case discussion was recorded in the trauma care quality register and was presented as a binary variable, “Yes -
At least one OFI identified” and “No - No OFI identified”.

@@

Certain continuous variables were retained in their continuous form for analysis, but others were categorised into
clinically meaningful groups to facilitate descriptive statistics and subgroup analyses.  The continuous variables
include…  
And those who were categorised were....

Furthermore, the categorical variables include…
<!-- Choose variables, not sure which ones to chose--> 
<!-- MGW: Start with the ones in Husseins and Elins papers-->

Statistical methods
----------------

For continuous variables, medians and interquartile ranges (25th-75th  percentiles) are used. Categorical variables
are presented using the number of patients and their corresponding percentages of the cohort. <!-- If multivariable regression analysis will be used, how will we handle missing data from other variables for the same patient? -->

<!-- MGW: Regardless of the analysis method, we need to handle missing data somehow. You can use complete case analysis, which means that you only include cases with complete data for all variables of interest. -->

Logistic regression was used to examine the association between patient and process-related factors and OFI.
Bivariable logistic regression to assess unadjusted associations between patient and process factors and OFI was
initially performed. Variables of clinical relevance and those significant in bivariable analyses were then entered
into a multivariable logistic regression model to estimate adjusted associations. <!-- MGW: I think you should decide which ones to include in the multivariable model from the start, regardless of their significance in the bivariable analysis --> Results are reported as Odds Ratio
(OR) with 95% confidence interval (CI), and a p-value of <0.05 was used as the threshold for statistical
significance. To ensure objectivity, all statistical analyses were initially performed on synthetic data before being
applied to the combined trauma registry and trauma care quality database. Statistical analyses were conducted in R
version x.

<!-- *Refer to the appropriate reporting guideline for details. If you are
developing, updating or validating a clinical prediction model then
use
[TRIPOD](https://www.equator-network.org/reporting-guidelines/tripod-statement/). If
you are conducting an observational study, for example a cohort or
case control study in which you assess associations between some
exposure and an outcome then use
[STROBE](https://www.equator-network.org/reporting-guidelines/strobe/).* -->

@@

*This is part B) in the project plan [sv: projektredogörelse]. *

*Save the image generated from the Gantt chard spreadsheet into the assets folder, rename it to project-gantt.png*

![**The Gantt chart will be displayed here when you place it in the assets folder under "project-gantt.png"**](assets/project-gantt.png)

Back-up Plan
============

*This is part C) in the project plan [sv: projektredogörelse]*

>>>>

---
title: "project-plan"
date: today 
bibliography: bibliography.bib
csl: vancouver.csl
format: html
---
Introduction
============
Injury has reached epidemic proportions worldwide, disproportionately affecting the most productive and the youngest
members of society. Trauma, often synonymously used with serious injury, is defined as a physical injury and the
body's associated response [@Cao]. Furthermore, trauma remains a leading cause of death and disability
worldwide, where it is the foremost cause of death among people under the age of 45 and ranks as the fourth leading
cause of death across all ages [@Maerz2009].

@@

Among trauma-related conditions, traumatic brain injury (TBI) is a major cause of death and disability worldwide and
constitutes a significant public health burden [@Guan2023]. TBI is the result of an external force causing
neurological impairments that may be either transient or permanent [@Capizzi2020]. Furthermore, TBI should not be
regarded as a single event but rather as a dynamic process. The initial trauma often triggers a secondary
pathophysiological cascade, including ischemia, oedema, and metabolic disturbances, that develop over time and
critically influence long-term outcomes [@Robba2025]. As such, patients with TBI often require admission to the
intensive care unit (ICU).

@@

Initial management of TBI spans prehospital and early in-hospital phases and centres on rapid physiological
stabilisation and diagnostics in order for secondary injuries to be limited. Protection of the airway in patients
with impaired consciousness, adequate oxygenation and normocapnia, and fast correction of hypotension are all
prioritised in the initial phase of management as hypoxia and low blood pressures are associated with an increased
mortality  [@Carney2017; @Stocchetti2017; @Chesnut1993; @Spaite2017]. Non-contrast head CT should be performed
without delay to identify surgically remediable lesions. Depending on mechanism and physiology, whole-body CT may be
used selectively, as randomised data show no mortality advantage over targeted imaging, while observational work
underscores the benefit of immediate CT pathways for severely injured patients [@Sierink2016; @Yamamoto2023].
Haemostatic strategies in the early stages, such as administering tranexamic acid within three hours of injury, can
reduce mortality [@Crash-32019]. After this initial phase, depending on the patient's condition, management
transitions to other areas of the hospital, where prevention of secondary systemic and intracranial insults becomes
the primary focus.

The management of TBI in the ICU is particularly challenging. Given the heterogeneity and dynamic nature of the
condition, care must prioritise the prevention and limitation of secondary systemic threats such as hypoxia,
hypercapnia, arterial hypotension, hyponatraemia and pyrexia. Equal attention must also be directed toward secondary
intracranial threats, including expanding haematomas, cerebral contusions and rises in intracranial pressure
[@Robba2025; @Stocchetti2017; @Slot2025]. Although these complexities are often anticipated, they render ICU care
especially vulnerable to deviations from optimal practice, thereby posing risks to both patient safety and overall
quality of care.

The intensive care of patients with TBI has been addressed in both guidelines and reviews. Multiple recent reviews
and guideline updates consistently refer to the Brain Trauma Foundation (BTF) Fourth Edition, which remains a central
reference for modern management [@Robba2025; @Stocchetti2017; @Slot2025; @Wiles2022; @Dash2018]. The BTF advises
treatment strategies to lower ICP and how to maintain cerebral perfusion pressure (CPP) within optimal ranges. It
also discusses what practices to avoid, such as prophylactic hyperventilation within the first 24 hours and
corticosteroid usage.

@@

Despite the presence of systems and guidelines in place for these patients, trauma care remains vulnerable to errors,
deviations from established practice, and treatment delays, all of which may negatively influence outcomes <!-- CP: Reference-->. In response, the American College of Surgeons developed the Trauma Quality Improvement Program (TQIP),
a risk-adjusted benchmarking system designed to assess the quality of trauma care <!-- MGW: I would mention ACS TQIP
as one example of a TQIP and refer to the WHO guidelines as more general. I suggest that you start by introducing the
concept of TQIP by referring to the WHO guidelines and then mention ACS TQIP as one example -->. Through audit filters and peer review, it identifies opportunities for improvement (OFI) and ultimately reduces trauma-related mortality [@Al-thani2023; @Shafi2009]. The WHO, however, insist that the trauma quality improvement programmes must be a continuous, multidisciplinary process. This entails audit filters, structured morbidity and mortality and preventable death case reviews in order to find opportunities for improvement (OFI). They also discuss that the formulation of corrective strategies or action plans in response to identified OFI:s is essential to trauma quality improvement programmes. Furthermore, the WHO emphasises that TQIP must ensure that corrective measures addressing identified OFI result in tangible and verifiable improvements in outcomes, thereby confirming that deficiencies are not only recognised but also effectively resolved [@world_health_organization_guidelines_2009].

Previous studies and reviews have discussed preventable trauma deaths where recurring patterns occur, namely, delays
in imaging and care, missed injuries, deviations in care regarding airway-related care, haemorrhage control, and
fluid management [@Sanddal2011; @Gruen2006; @Alao2024; @Teixeira2007; @Mabry2024]. Importantly, some of these
findings were observed in high-performing level I trauma centres, underscoring that even mature systems remain
vulnerable to error. Furthermore, injuries to the central nervous system, like TBI, account for a substantial
proportion of trauma-related deaths, as demonstrated in analyses from large registries such as the Arkansas Trauma
Registry [@Mabry2024]. Yet, while previous research has described OFI in trauma resuscitation and in general ICU
trauma cohorts, relatively little is known about the prevalence, nature, and determinants of OFI specifically among
patients with TBI.

Rationale
------------

TBI remains a challenge within modern trauma systems, not only due to its contribution to mortality but also due to
the magnifying impact on the outcome because of small process deviations. Delays in imaging, inconsistent airway
management, and lapses in haemodynamic measures are all examples that can result in irreversible damage. Although
TQIP have been developed to identify and correct such lapses, with the WHO stressing the importance of linking OFIs
to corrective actions and verification of bettered outcomes, most research has been done on mixed trauma cohorts.
This, on the other hand, leaves TBI-specific vulnerabilities undercharacterised. Recent work by *Szolnoky et al.*
suggests that, despite overall declines in OFI, patients with TBI remain at risk of preventable shortcomings,
underscoring the need for TBI-specific quality improvement efforts [@Szolnoky2025]. <!--CP: Therefore we hypothesis ze that is possible to identify new OFis related to TBI?? 
CP: See if something like that can be a hypothesis -->

Aim
=====
This study aims to identify and describe OFI in the care of adult trauma patients with TBI, and to examine patient- and process-related factors associated with their occurrence. <!-- CP: You need to phrase these aims in a more concrete way: this what this course demands..see also next comment E.g. (but for your supervisor to help with) Identify specific OFI by doing what? I mean , an identification require that certain parameters will be used and confirmed
If too long M&M will describe these parameters)” -->

<!-- CP: Pls describe this aim in a Yes/No question e..g determine whether the identified OFI are associated with patient and process related factors. Ask for your supervisor to help revising -->

<!--CP: Can you explain more concretely What this means? Occurrence of OPI?-->

@@

Study Design
-----------
A single-centre, registry-based retrospective cohort study was conducted at Karolinska University Hospital in Solna
(NKS). The trauma registry for NKS is a comprehensive trauma registry that is part of the Swedish National Trauma
Registry, SweTrau, and is compliant with the Utstein template for uniform reporting after trauma [@Ringdal2008]. As
such, the registry records patient demographics, prehospital and emergency department physiology, times of key
interventions such as imaging, injury severity scores, and in-hospital outcomes.

The Swedish Trauma Registry includes patients of all ages who have sustained serious injuries from traumatic events
such as falls, traffic incidents, or other external violence. Patients have been included if a trauma team activation
has occurred, regardless of injury severity, or if they have a New Injury Severity Score (NISS) greater than fifteen
without trauma team activation. Patients transferred to a hospital within seven days of the traumatic event with a
NISS greater than fifteen are also included. Exclusion criteria are cases where the only traumatic condition is
chronic subdural haematoma, trauma alarms without an underlying traumatic event, patients with protected identity,
and patients declared dead on arrival. Since these inclusion and exclusion criteria have changed over time, patients
in this study will be filtered to ensure they fulfil the criteria relevant to the study period [@Holmberg_2023].

@@

Setting
---------
Karolinska University Hospital in Solna houses Trauma Centre Karolinska, Sweden’s largest provider of advanced trauma
care for both adults and children. According to the definition of the American College of Surgeons, NKS serves as the
designated Level 1 trauma centre for the greater metropolitan area of Stockholm. This designation requires
around-the-clock availability of key services, including neurosurgery, general surgery, orthopaedic surgery,
anaesthesia and intensive care, as well as diagnostic and interventional radiology and consultants from other
relevant specialities [@American_college_of_surgeons_resources_2022; @Granstrom2018]. Additionally, the ICU units in
NKS are all recognised as Level 3 units according to the Swedish Society of Anaesthesiology and Intensive Care
Medicine [@Swedish_society_of_anaesthesiology_and_intensive_care_medicine_riktlinjer_2024].

Furthermore, the trauma registry covers trauma patients admitted between 2012 and 2023. The trauma care quality
database is a subset of this registry and contains patients who have been selected for review between 2014 and 2023.
Screening for potential opportunities for improvement was implemented in 2013 for selected patient cases. Initially,
case selection was based on a limited number of audit filters and the judgment of a small group of clinicians
directly involved in trauma care. Since 2017, a standardised set of audit filters has been consistently applied, and
all trauma patients admitted to the hospital are screened for potential opportunities for improvement. The process
has expanded to include a broader group of senior clinicians and nursing staff, ensuring a systematic and
comprehensive review. Patients flagged through the audit filters, and therefore have potential OFI, undergo
discussion in structured multidisciplinary morbidity and mortality conferences, where representatives from relevant
specialties and professions involved in trauma care are involved.  The discussion focuses on establishing whether
there were any preventable events or deviations from standards of practice. The final classification of whether an
opportunity for improvement is present is determined by consensus among the peer-review group. As such, subjectivity
is minimised and ensures that decisions reflect a collective judgement. Whether or not an OFI is identified, the
decision of the review is recorded in the trauma care quality database, ensuring systematic documentation of positive
findings and cases without deviations.

Participants
------------
Included in the study were all adult patients with TBI recorded in the Karolinska trauma registry between 2012 and
2023 who also had been screened for opportunity for improvements. Furthermore, TBI was defined as a Glasgow Coma
Scale (GCS) score of ≤8 combined with an Abbreviated Injury Scale (AIS) ≥3 for the head region. Patients were
eligible irrespective of level of care and were not restricted to intensive care admissions.

Patients younger than 18 years were excluded, as their trauma follows separate clinical care systems and quality
review processes that differ from those for adults. Also excluded were patients declared dead on arrival, since
assessment of opportunities for improvement in these cases is not relevant to the aim of this study.

@@

The primary outcome of this study was the presence of one or more OFIs during the management <!-- CP: You probably have all needed info in this text. However, Pls describe this section in relation to Aim 1 and 2. The text here needs to be stringently matched with the two aims-->
of patients with TBI. As mentioned, OFIs were adjudicated after a screening process with audit filters and a
subsequent review in multidisciplinary morbidity and mortality conferences. Following the peer review, the result of
each case discussion was recorded in the trauma care quality register and was presented as a binary variable, “Yes -
At least one OFI identified” and “No - No OFI identified”.

@@

The categorical variables include sex, on-call times defined as hours outside of 08:00 - 17:00, or Saturday and
Sunday, hospital care level ordered from the emergency department to the highest being the ICU, intubation status,
mechanism of injury, ASA class, duration of mechanical ventilation and ICU length stay. The latter two are
categorised into different groups that represent a certain number of days.

The continuous variables include age, Injury Severity Score (ISS), Revised Trauma Score (RTS), systolic blood
pressure (SBP), respiratory rate (RR) and time to first CT in minutes. The Revised Trauma Score was calculated using
GCS, systolic blood pressure, and respiratory rate, each coded on a scale from 0 to 4 and combined with published
coefficients [@Champion1989]. Furthermore, certain continuous variables were retained in their continuous form for
analysis, but others were categorised into clinically meaningful groups to facilitate descriptive statistics and
subgroup analyses.

Statistical methods. <!-- CP: Pls recall aim 1 and aim 2 and develop the below text in relation to these 2 aims: see criteria-->
----------------

For continuous variables, medians and interquartile ranges (25th-75th  percentiles) are used. Categorical variables
are presented using the number of patients and their corresponding percentages of the cohort. Complete case analysis
will be used; as such, only cases with complete data for all variables of interest will be used.

Logistic regression was used to examine the association between patient and process-related factors and OFI.
Bivariable logistic regression to assess unadjusted associations between patient and process factors and OFI was
initially performed. To estimate adjusted associations, these variables were entered into a multivariable logistic
regression model. Results are reported as Odds Ratio (OR) with 95% confidence interval (CI), and a p-value of <0.05
was used as the threshold for statistical significance. To ensure objectivity, all statistical analyses were
initially performed on synthetic data before being applied to the combined trauma registry and trauma care quality
database. Statistical analyses were conducted in R.

Ethical considerations
=======

This is a register-based study with no intervention and thus entailed no risk of burden nor physical harm or contact
with patients. Rather, it involves secondary use of routinely collected clinical data. The principal ethical risk was
the risk of privacy and integrity intrusion, especially during the linkage of the trauma registry and the local
trauma care quality database before pseudonymization. Also, all results are reported on an aggregate level to
minimise the possibility of reidentification. In line with regular practice for registry studies based on already
collected data, informed consent was not obtained from participants. The risk of privacy intrusion was judged to
increase if individual research subjects were to be contacted.

The risk of integrity breach was weighed against the potential benefits the project can provide regarding health and
was judged favourable. Although there is no direct benefit for the individuals whose data was used, the societal
value is substantial because identifying opportunities for improvement in traumatic brain injury care can inform
corrective actions, enhance patient safety, and reduce morbidity and mortality. The Swedish Ethical Review Authority
granted ethical approval with the DNRs being: 2021-02541, 2021-03531 and 2023-02975-02.

@@

Data collection has already been done using two established registers at Karolinska University Hospital, Solna. These
are the trauma registry (SweTrau) with approximately 14,000 patients between 2012–2023, and the trauma care quality
database, which contains around 8,000 patients between 2014–2023. The latter includes the nominable variable of Yes
or No OFI identified during multidisciplinary morbidity and mortality conferences.

Variables: Patient and process factors are analysed

-	Categorical: sex, arrival outside office hours, hospital care level, intubation status, mechanism of injury, ASA
class, ventilation duration, and ICU length of stay.

-	Continuous: age, Injury Severity Score, Revised Trauma Score, systolic blood pressure, respiratory rate, and time to CT. Some continuous variables will be categorised.

Analysis:

-	Results will be presented in tables and figures with for eg. Means, interquartile ranges, counts and percentages.

-	Associations between factors and OFI will be examined using bivariable and multivariable logistic regression

This will be presented as OR with 95% CI. Analyses will be complete-case and conducted in R.

![**Gantt timeline for the project**](assets/project-gantt.png)
Back-up Plan
============

If unforeseen issues arise during the project, and as data has already been collected, the project will be adjusted by doing the following:

- **Reduce the dataset scope**  
  Instead of analysing all patients with TBI in the register between 2012–2023, the reserve plan would focus on a shorter time period and therefore reduce the number of patients analysed.

- **Reduce the variable set**  
  If analysis of the full range of patient- and process-level factors proves too difficult to do within the timeframe, the analysis will prioritise some variables over others and reduce the number of variables.

- **Alternative analysis**  
  If multivariable regression is not possible or proves too difficult, the reserve plan is to perform simpler analysis, such as simple/bivariable analysis, therefore only presenting unadjusted associations.
